Introduction to Structural Causal Models

### 2021-05-21 presentation_video.Rmd
minoor
presentation

@@

- Nonparametric SEM 
- System of equations with functions
- **Assignments** ':=' (non-symmetric) 
- algebraic equation '=' (symmetry)
- **Error in Regression:** Omittable outside factor 
- **Error in SCM:** Latent (influential) factor that is pivotal for the model

@@

source: @peters_elements_2017

# Assumptions in Causality (1)

@@

# Assumptions under which causal inference is possible (2)

# SCM Applications:

- Flexible simulations for higher order problems (intervention, counterfactual) based on elementary noise variables
- Graphical visualization via directed acyclic graph
- Example: SCM for fairness in dynamical system [@creager] : credit loan approval, time allocation, college admission

@@

# Prediction

- ML, BN and regression are at the lowest level in the causal hierarchy
- Prediction methods demand the least information and depend on association alone
- Association-based methods ignore external changes outside of our data
- Intervention distribution needed for higher level information

@@

- missing data problem in PO framework

Process is described as follows:

@@

@@

- Nodes -> Variables (endogenous/exogenous)
- Edges -> relationship (equations)
- Parents/Ancestors/Descendants
- No need to specify exact parametric shape
- Highlight colliders 
- Estimation back door criterion 
- Test theoretical model structure via causal algorithms to detect structure in data (IC/PC Algo.) 
- More information: [@morgan_winship_2014]

@@

```{tikz,fig.align="center",fig.cap="Probabilistic Model", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\node[my node] (A){A};
\node[my node,right =1 cm of A](C){C};
\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
\draw[->] (A) -- (T);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

@@

```{tikz,fig.cap="Structural Causal Model", fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

@@

# Extensions

- Mediation Analysis 
- PO-Framework
- Causal Algorithms
- IV-Estimation
- Causal Constraints Model 
- Causal Trees

>>>>

# Seminar Topics

@@

- Flexible simulations for higher order problems (intervention, counterfactual) based on elementary noise variables
- Graphical visualization via directed acyclic graph
- Nonparametric SEM 
- System of equations with functions
- **Assignments** ':=' (non-symmetric) 
- algebraic equation '=' (symmetry)
- **Error in Regression:** Omittable outside factor 
- **Error in SCM:** Latent (influential) factor that is pivotal for the model

@@

Source: @peters_elements_2017

# Assumptions (1)

@@

@@

@@

- Missing data problem in PO framework 
- SCM Process is described as follows:

@@

# Assumptions

@@

- Nodes -> Variables (endogenous/exogenous)
- Edges -> relationship (equations)
- Parents/Ancestors/Descendants
- No need to specify exact parametric shape
- Focus on acyclical structures (DAG)
- Estimation back door criterion 
- Test theoretical model structure via causal algorithms to detect structure in data (IC/PC Algo.) 
- More information: [@morgan_winship_2014]

# Undirected/Semidirected Graph

```{tikz, well, fig.cap ="Undirected graph", fig.align="center", echo=F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{line} = [thick,-,>=stealth, line width=0.5pt]
\node[my node] (A){A};
\node[my node,right of=A](C){C};
\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (B) {B};
\draw[line] (A) -- (B);
\draw[line] (C) -- (A);
\draw[line] (B) -- (C);
\end{tikzpicture}
```

```{tikz, wela, fig.cap ="Semi-directed graph", fig.align="center", echo=F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{line} = [thick,-,>=stealth, line width=0.5pt]
\node[my node] (A){A};
\node[my node,right of=A](C){C};
\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (B) {B};
\draw[line] (A) -- (B);
\draw[line] (C) -- (A);
\draw[->] (B) -- (C);
\end{tikzpicture}
```

@@

```{tikz,fig.align="center",fig.cap="Probabilistic Model", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\node[my node] (A){A};
\node[my node,right =1 cm of A](C){C};
\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (E) {E};
\draw[->] (A) -- (E);
\draw[->] (A) -- (C);
\draw[->] (E) -- (C);
\end{tikzpicture}
```

@@

```{tikz,fig.cap="Structural Causal Model", fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (E) {E};
\node[my node,right = 1 cm of E, fill=gray!30](C){C};
\node[latent,left = 1cm of E](UE){$U_E$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (E);
\draw[->] (UA) -- (A);
\draw[->] (UE) -- (E);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (E) -- (C);
\end{tikzpicture}
```

@@

# Does and Do nots

**If you...**

* ...only have conditional probabilities 
* ...are only interested in association 
* ...have very limited knowledge of your actual underlying causal structure

$\rightarrow$ you should probably go for a probabilistic model, association based model or a model based on conditional probabilities

**But if you**

* ... have a good understanding of underlying latent factors 
* ... have a lot of information on your underlying relationship in your data 
* ... are dealing with high stake settings 
* ... want to simulate different interventions before implementing them

$\rightarrow$ you should probably go for a SCM
